Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2019-03-04 paper/paper.tex
updated contact
updated sentence in abstract

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.5in,footskip=0.75in]{geometry}

@@

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\definecolor{linkblue}{RGB}{0,76,153}
\usepackage[colorlinks=true,linkcolor=linkblue,citecolor=linkblue,urlcolor=linkblue]{hyperref}
\usepackage{nameref}

@@

% added by Jesse Bloom for full-width figures
\newenvironment{fullwidth}{%
  \begin{adjustwidth}{-2.05in}{}
}{\end{adjustwidth}}

% Added by Jesse Bloom for figures
\renewcommand{\floatpagefraction}{0.99}
\renewcommand{\textfraction}{0.01}
\setcounter{topnumber}{8}
\setcounter{bottomnumber}{8}
\setcounter{totalnumber}{8}

@@

\newfloat{suppfig}{thp}{lofsupfile}
\floatname{suppfig}{Figure}
\renewcommand{\thesuppfig}{S\arabic{suppfig}}

\renewcommand{\figureautorefname}{Figure}
\newcommand{\FIG}[1]{\autoref{fig:#1}}

@@

\newcommand{\suppfigautorefname}{Figure}
\newcommand{\SUPPFIG}[1]{\autoref{suppfig:#1}}

@@

% Text layout
%\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.5in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Figure}

% Harvard-style citations
\usepackage{natbib}
\bibliographystyle{cell}

@@

% section header formats
\usepackage{sectsty}
\definecolor{darkred}{RGB}{140,0,0}
\sectionfont{\normalsize \color{darkred}}
\subsectionfont{\normalsize \color{darkred}}
\subsubsectionfont{\normalsize \color{darkred}}
\paragraphfont{\normalsize \color{darkred}}

@@

\providecommand{\keywords}[1]{\noindent \textbf{Keywords:} #1}

\newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\LARGE
\textbf{Single-cell virus sequencing of influenza infections that trigger innate immunity}}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
{\bf
Alistair B. Russell\textsuperscript{1},
Jacob R. Kowalsky\textsuperscript{2},
Jesse D. Bloom\textsuperscript{1,2,3*}
}
\\
\bigskip
{\small
$^1$Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center
\\
$^2$Department of Genome Sciences, University of Washington
\\
$^3$Howard Hughes Medical Institute
\\
Seattle, WA  98109, USA
\\
*Correspondence: jbloom@fredhutch.org
}
\bigskip

\end{flushleft}

\section*{SUMMARY}
{\bf
The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
Here we develop a new approach to assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by determining both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Most cells are infected by virions with defects such as amino-acid mutations, internal deletions, or failure to express a gene.
We identify instances of each type of defect that increase the likelihood that a cell activates an innate-immune response.
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even when a cell is infected by a virion that expresses unmutated copies of all genes.
Our work shows that viral genetic variation substantially contributes to but does not fully explain the heterogeneity in single influenza-infected cells.
}

\vspace{0.2in}
\keywords{PacBio, single-cell RNA-seq, influenza, interferon, NS1, stochastic}

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{INTRODUCTION}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and produce interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.

Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\citep{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It is unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
But for influenza, a third possible contributor looms large: viral genetic diversity.
Because influenza has a high mutation rate, individual virions often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
None of these techniques reliably reveal if there are mutations in the virion infecting any given single cell.

Here we develop a new approach to determine both the transcriptome and full sequences of all viral genes in single influenza-infected cells.
To do this, we perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
Two-thirds of cells are infected by virions with a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by unmutated virions having a tighter distribution of viral transcriptional burden.
We identify several types of viral defects that increase IFN induction.
However, viral genetic variation does not fully explain the heterogeneity, and even unmutated virions sometimes induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

\section*{RESULTS}

\subsection*{A system to identify and enrich rare IFN+ cells}

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.34\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large B} \includegraphics[width=0.22\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
{\bf Reporter cells to identify and enrich infections that activate IFN expression.}
{\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\SUPPFILE{plasmids}).
In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
{\bf (B)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{reportervalidation}C for full data).
The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}
\end{figure}
%%% end IFNrare figure %%%

A challenge in studying IFN induction by influenza virus is its rareness at the level of single cells.
Studies using reporters~\citep{killip2017single,kallfass2013visualizing} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that at early timepoints IFN induction occurs in $<$1\% of influenza-infected cells.
To efficiently identify and enrich these rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein ~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
For the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells~\citep{russell2018extreme}.

@@

\caption{
{\bf Approach for combined transcriptomics and virus sequencing of single influenza-infected cells that express IFN.}
{\bf (A)}
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\SUPPFIG{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
{\bf (B)}
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
{\bf (C)}
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
{\bf (D)}
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
{\bf (E)}
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

@@

To determine if mutations or other defects in the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{plasmids}).
These viral barcodes allow us to identify co-infections~\citep{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter (\SUPPFIG{MACS}).
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

We processed the cells on a commercially available platform~\citep{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\citep{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\citep{robertson1981polyadenylation} in each of the eight viral gene segments.

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\citep[CCSs;][]{travers2010flexible}.
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).

@@

We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
To remove some of these multiplets along with low-quality droplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts as is commonly done in analysis of single-cell RNA-seq data~\citep{haque2017practical}.
After this filtering, we retained 1,490 human cells for further analysis (\FIG{transcriptomics}B).

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is from lysed cells that release ambient viral mRNA that is captured in droplets prior to reverse-transcription.
We tested whether each cell contained more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected by virus (\FIG{transcriptomics}C).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
The distribution of the amount of viral mRNA across infected cells is in the inset in \FIG{transcriptomics}C.
As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

>>>>

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

@@

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

@@

% added by Jesse Bloom for full-width figures
\newenvironment{fullwidth}{%
  \begin{adjustwidth}{-2.3in}{}
}{\end{adjustwidth}}

% Added by Jesse Bloom for figures
\renewcommand{\floatpagefraction}{.9}
\setcounter{topnumber}{8}
\setcounter{bottomnumber}{8}
\setcounter{totalnumber}{8}

@@

\newfloat{suppfig}{thp}{lofsupfile}
\floatname{suppfig}{Fig}
\renewcommand{\thesuppfig}{S\arabic{suppfig}}

\renewcommand{\figureautorefname}{Fig}
\newcommand{\FIG}[1]{\autoref{fig:#1}}

@@

\newcommand{\suppfigautorefname}{Fig}
\newcommand{\SUPPFIG}[1]{\autoref{suppfig:#1}}

@@

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

@@

@@

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Single-cell virus sequencing of influenza infections that trigger innate immunity} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Alistair B. Russell\textsuperscript{1},
Jacob R. Kowalsky\textsuperscript{2},
Jesse D. Bloom\textsuperscript{1,2,3*}
\\
\bigskip
\textbf{1} Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States
\\
\textbf{2} Department of Genome Sciences, University of Washington, Seattle, United States
\\
\textbf{3} Howard Hughes Medical Institute, Seattle, United States
\\
\bigskip

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* jbloom@fredhutch.org

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Infections that activate an innate-immune response in single cells are often initiated by virions with defects that include amino-acid mutations, internal deletions, and failure to express key genes.
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
Our work shows that the genetic variation present in influenza virus populations contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
But so far, it has been impossible to fully characterize the genetic properties of the virions infecting any given cell, since the conventional techniques of flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not its sequence.
Here we develop a new approach that uses long-read PacBio sequencing to determine the full sequences of the virions infecting individual cells.
Most cells are infected by virions that have a mutation or some other defect.
This viral genetic variation explains a substantial amount of the cell-to-cell variability in how much viral mRNA is produced in infected cells, and helps determine whether a cell activates an innate-immune response.
However, our results also show that viral genetic defects do not completely explain cell-to-cell variation in the outcome of infection.
Overall, this study provides the first complete picture of how viral mutations affect the outcome of infection in single cells.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and begin producing interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\cite{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\cite{la2007question, iwasaki2014innate}.

Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It is unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.

Here we develop a new approach to measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
Even though we used a ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
Viral defects are especially common in IFN+ cells, and we identify at least four types of defects that cause increased IFN induction.
However, viral genetic variation does not fully explain the heterogeneity, as even wild-type virions sometimes induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

\section*{Results}

\subsection*{A system to identify and enrich for rare IFN+ cells}

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large B} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
{\bf Reporter cells to identify and enrich infections that activate IFN expression.}
{\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
{\bf (B)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
The plot also shows uninfected cells, and cells infected with Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}
\end{figure}
%%% end IFNrare figure %%%

A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).

To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

We generated a stock of ``wild-type'' A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.

@@

\caption{
{\bf Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.}
{\bf (A)}
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\SUPPFIG{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
{\bf (B)}
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
{\bf (C)}
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
{\bf (D)}
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
{\bf (E)}
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

@@

To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

We processed the cells on a commercially available platform~\cite{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
Because influenza virus mRNAs are polyadenylated~\cite{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\cite{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\cite{robertson1981polyadenylation} in each of the eight viral gene segments.

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\cite{travers2010flexible}).
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).

@@

We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
